---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simoptim

<!-- badges: start -->
![R-CMD-check](https://github.com/audrey-b/simoptim/workflows/R-CMD-check/badge.svg)
<!-- badges: end -->

`simoptim` is an R package for optimal allocation of sampling effort in
Bayesian simulation studies with MCMC. Given estimates of two variance
components — between-dataset variance and within-dataset posterior variance
— it computes the optimal number of simulation datasets `r` and effective
posterior samples per dataset `m` that minimise total computational cost
subject to a precision target.

> **Note:** This package is under active development. The API may change.

## Installation

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("audrey-b/simoptim")
```

## Background

In a Bayesian simulation study, the precision of the estimator is affected
by two sources of Monte Carlo error:

- **Between-dataset variance** ($\sigma_1^2 / r$): variability across the
  `r` simulated datasets.
- **Within-dataset variance** ($\sigma_2^2 / (r \cdot m)$): variability due
  to running MCMC for only `m` effective samples per dataset.

`simoptim` implements the methodology of Beliveau et al. (in preparation)
to choose `r` and `m` optimally given a precision target $p = \Delta /
\sigma_1$ and the ratio of burn-in time to time per effective sample.

## Usage

The typical workflow has four steps.

### Step 1: Run a pilot study

Run MCMC on a small number of pilot datasets and collect the results as a
list of `mcmc.list` objects. Here we simulate pilot data for illustration.

```{r}
library(simoptim)
library(coda)
set.seed(1234)

simulate_mcmc_list <- function(n_iter = 500, n_burnin = 100, n_chains = 3) {
  # Draw the estimator thetahat ~ N(0, 1)
  thetahat <- rnorm(1, mean = 0, sd = 1)
  
  # Posterior is N(thetahat, 1), same variance
  chains <- lapply(1:n_chains, function(i) {
    coda::mcmc(
      matrix(rnorm(n_iter, mean = thetahat, sd = 1), ncol = 1,
             dimnames = list(NULL, "theta")),
      start = n_burnin + 1
    )
  })
  coda::mcmc.list(chains)
}

analyses_results <- lapply(1:8, function(i) simulate_mcmc_list())
```

### Step 2: Summarise posterior samples

```{r}
summaries <- lapply(analyses_results, summarize_mcmc, parameter = "theta")
means  <- sapply(summaries, `[[`, "mean")
mcmcses <- sapply(summaries, `[[`, "mcmcse")
vars   <- sapply(summaries, `[[`, "var")
```

### Step 3: Estimate variance components

```{r}
var_ests <- estimate_variances(means, mcmcses, vars)
var_ests
```

### Step 4: Compute optimal allocation

```{r}
optimize_allocation(burnin_time = 10,
      eff_sample_time = 1,
      var_between = var_ests$var_between,
      var_within = var_ests$var_within,
      min_m = 20,
      p = 0.05)
```

If only ratios are available, `optimize_allocation_from_ratios()` can be
used instead and returns the same result:

```{r}
variance_ratio <- var_ests$var_between / var_ests$var_within
optimize_allocation_from_ratios(time_ratio = 10,
        variance_ratio  = variance_ratio,
        min_m      = 20,
        p          = 0.05)
```

See `vignette("simoptim")` for a full worked example.

## Citation

Beliveau A., Stringer, A. and Tan C. (in preparation). Optimal allocation of sampling effort in Bayesian simulation studies with MCMC.

## License

MIT
