---
title: "Getting Started with simoptim"
author: "Audrey Beliveau"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with simoptim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

In a Bayesian simulation study, two sources of Monte Carlo error affect the
precision of your estimator:

- **Between-dataset variance** ($\sigma_1^2 / r$): variability across the `r` simulated datasets.
- **Within-dataset variance** ($\sigma_2^2 / (r \cdot m)$): variability due to running MCMC for only `m` effective samples per dataset.

`simoptim` helps you choose `r` and `m` to meet a precision target while
minimising total compute time.

```{r setup}
library(simoptim)
library(coda)
set.seed(1234)
```

## Step 1: Generate pilot MCMC results

In practice you would supply real MCMC output. Here, we simulate
`mcmc.list` objects directly from a normal distribution to illustrate
the workflow.

We assume that, under large samples, the estimator $\hat{\theta}$ is
approximately normally distributed:
$$\hat{\theta} \sim N(\theta, \sigma_1^2)$$
with true value $\theta = 0$ and $\sigma_1^2 = 1$.

We further assume the posterior distribution is normal with mean
$\hat{\theta}$ and the same variance:
$$\theta \mid \hat{\theta} \sim N(\hat{\theta}, \sigma_2^2)$$
with $\sigma_2^2 = 1$.

Under this setup, $\sigma_1^2 = \sigma_2^2 = 1$, so the variance ratio
$\sigma_1^2 / \sigma_2^2 = 1$.

```{r}
simulate_mcmc_list <- function(n_iter = 500, n_burnin = 100, n_chains = 3) {
  # Draw the estimator thetahat ~ N(0, 1)
  thetahat <- rnorm(1, mean = 0, sd = 1)
  
  # Posterior is N(thetahat, 1), same variance
  chains <- lapply(1:n_chains, function(i) {
    coda::mcmc(
      matrix(rnorm(n_iter, mean = thetahat, sd = 1), ncol = 1,
             dimnames = list(NULL, "theta")),
      start = n_burnin + 1
    )
  })
  coda::mcmc.list(chains)
}

analyses_results <- lapply(1:8, function(i) simulate_mcmc_list())
```

## Step 2: Summarise posterior samples

`summarize_mcmc()` extracts the posterior mean, MCMCSE, and variance for a
given parameter from each `mcmc.list`.

```{r}
summaries <- lapply(analyses_results, summarize_mcmc, parameter = "theta")

means  <- sapply(summaries, `[[`, "mean")
mcmcses <- sapply(summaries, `[[`, "mcmcse")
vars   <- sapply(summaries, `[[`, "var")
```

## Step 3: Estimate variance components

`estimate_variances()` uses the pilot summaries to estimate \eqn{\sigma_1^2}
(between-dataset) and \eqn{\sigma_2^2} (within-dataset).

```{r}
var_ests <- estimate_variances(means, mcmcses, vars)
var_ests
```

## Step 4: Compute optimal allocation

`optimize_allocation()` takes individual values of burn-in time, time per effective sample,
and the two variance components, along with your precision target `p`, and
returns the optimal `r` and `m`.

```{r}
optimize_allocation(burnin_time = 10,
      eff_sample_time = 1,
      var_between = var_ests$var_between,
      var_within = var_ests$var_within,
      min_m = 20,
      p = 0.05)
```

The output gives the recommended number of simulation datasets (`r`) and
effective posterior samples per dataset (`m`) to achieve a standardised
precision of `p = 0.05`, assuming the burn-in period takes 10 times longer
than generating one effective posterior sample.

Alternatively, `optimize_allocation_from_ratios()` accepts the same inputs as ratios, since the
optimal allocation depends only on these ratios rather than on the
individual quantities themselves.

```{r}
variance_ratio <- var_ests$var_between / var_ests$var_within
optimize_allocation_from_ratios(time_ratio = 10,
        variance_ratio  = variance_ratio,
        min_m      = 20,
        p          = 0.05)
```

The two functions return the same result.

