---
title: "Getting Started with simoptim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with simoptim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

In a Bayesian simulation study, two sources of Monte Carlo error affect the
precision of your estimator:

- **Between-dataset variance** (\eqn{\sigma_1^2 / r}): variability across the `r` simulated datasets.
- **Within-dataset variance** (\eqn{\sigma_2^2} / (r \cdot m)): variability due to running MCMC for only `m` effective samples per dataset.

`simoptim` helps you choose `r` and `m` to meet a precision target while
minimising total compute time.

```{r setup}
library(simoptim)
library(coda)
set.seed(1234)
```

## Step 1: Generate pilot MCMC results

In practice you would supply real MCMC output. Here, we simulate
`mcmc.list` objects directly from a normal distribution to illustrate
the workflow.

```{r}
simulate_mcmc_list <- function(n_iter = 100, n_burnin = 1000, n_chains = 3,
                                mu_mean = 0, mu_sd = 1,
                                lambda_mean = 1, lambda_sd = 0.5) {
  chains <- lapply(1:n_chains, function(i) {
    mcmc(cbind(
      mu     = rnorm(n_iter, mean = mu_mean, sd = mu_sd),
      lambda = rnorm(n_iter, mean = lambda_mean, sd = lambda_sd)
    ), start = n_burnin + 1)
  })
  
  mcmc.list(chains)
}

# 8 pilot datasets and their fictional run times
analyses_results <- lapply(1:8, function(i) simulate_mcmc_list())
run_times <- rnorm(8, 20, 2)
```

## Step 2: Summarise posterior samples

`summarize_mcmc()` extracts the posterior mean, MCMCSE, and variance for a
given parameter from each `mcmc.list`.

```{r}
summaries <- lapply(analyses_results, summarize_mcmc, parameter = "mu")

means  <- sapply(summaries, `[[`, "mean")
mcmcses <- sapply(summaries, `[[`, "mcmcse")
vars   <- sapply(summaries, `[[`, "var")
```

## Step 3: Estimate variance components

`estimate_variances()` uses the pilot summaries to estimate \eqn{\sigma_1^2}
(between-dataset) and \eqn{\sigma_2^2} (within-dataset).

```{r}
var_ests <- estimate_variances(means, mcmcses, vars)
```

## Step 4: Compute optimal allocation

`optim_2()` takes the ratio of burn-in time to per-effective-sample time and
the variance ratio, along with your precision target `p`, and returns the
optimal `r` and `m`.

```{r}
var_ratio <- var_ests$estimate_var1/var_ests$estimate_var2
optim_2(10, 
       var_ratio,
       min_m = 20, 
       p = 0.05)
```
The output gives the recommended number of simulation datasets (`r`) and
effective posterior samples per dataset (`m`) to achieve a standardised
precision of `p = 0.05` assuming the burn-in period takes 10 times longer than simulating an effective posterior sample.
